---
title: "Shorebird Data Analysis: thought and data processing"
author: "Emily Hamel"
date: "April 26, 2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(moderndive)
library(modelr)
library(broom)
library(knitr)
```
```{r data, include=FALSE}
hightide <- read_csv("DATA/High.tide.analysis.csv")
disturbance <- read_csv("DATA/Immed.displ.analysis.csv")
hightide_README <- read_csv("DATA/README_for_High.tide.analysis.csv")
disturbance_README <- read_csv("DATA/README_for_Immed.displ.analysis.csv")
```

## Background

These data sets were taken from a study conducted in Northern Europe, specifically the Dutch Wadden Sea island of Vlieland. The purpose of this study was to understand the impact of both short and long interval disturbance to the energy budget of roosting Eurasian Oystercatchers (*Haematopus ostralegus*).
The data analysis in this document seeks to recreate this study's results, in order to develop a more comprehensive understanding of analyzing ecological, bird-related data sets - a skill that will be necessary for my future professional research.

The following code and graphical information outlines this attempt to recreate the study's analysis and the thought processes involved. 

### Relevant Literature
Linssen, H., van de Pol, M., Allen, A.M. et al. Disturbance increases high tide travel distance of a roosting shorebird but only marginally affects daily energy expenditure. Avian Res 10, 31 (2019). https://doi.org/10.1186/s40657-019-0171-8

## Initial Data Analysis 

Following the 

