---
title: "Shorebird Data Analysis: thought and data processing"
author: "Emily Hamel"
date: "April 26, 2021"
pdf_document: 
    fig_caption: yes
    keep_tex: yes
header-includes: \usepackage{float}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(moderndive)
library(modelr)
library(broom)
library(readr)
library(knitr)
```
```{r data, include=FALSE}
hightide <- read_csv("DATA/High.tide.analysis.csv")
disturbance <- read_csv("DATA/Immed.displ.analysis.csv")
hightide_README <- read_csv("DATA/README_for_High.tide.analysis.csv")
disturbance_README <- read_csv("DATA/README_for_Immed.displ.analysis.csv")
```

## Background

These data sets were taken from a study conducted in Northern Europe, specifically the Dutch Wadden Sea island of Vlieland. The purpose of this study was to understand the impact of both short and long interval disturbance to the energy budget of roosting Eurasian Oystercatchers (*Haematopus ostralegus*).
The data analysis in this document seeks to recreate this study's results, in order to develop a more comprehensive understanding of analyzing ecological, bird-related data sets - a skill that will be necessary for my future professional research.

The following code and graphical information outlines this attempt to recreate the study's analysis and the thought processes involved. 

## Short-term Disturbance: initial data analysis 

The first piece of data to analyze in this study is the short-term impact on bird movement from different types of disturbance. In order to visualize the data I first created a plot (Figure 1) showing the immediate displacement resulting from different disturbance categories (namely: air force activity, non-anthropogenic events, and people walking past the roosting Oystercatchers). The "pre-disturbance" element of this plot seeks to show how disturbance related movements compare to the birds' activity prior to their displacement. The comparable plot in the literature is a box-plot, but I felt that the points provided a clearer picture of movement frequency and the related distance distribution - this choice, however, is simply a personal preference. 
From this data visualization it became important 

```{r immediate_dist, echo=FALSE, fig.pos = 'h', fig.cap="Figure 1. Depiction of immediate displacement resulting from different disturbance categories. The points are indicative of each displacement event of the 240 observations."}
ggplot(disturbance, aes(movement_category, distance)) + geom_point() + labs(x = "Movement Category", y = "Immediate Displacement Distance (km)", title = "Immediate Displacement as a Result of Disturbance")
```


### Relevant Literature
Linssen, H., van de Pol, M., Allen, A.M. et al. Disturbance increases high tide travel distance of a roosting shorebird but only marginally affects daily energy expenditure. Avian Res 10, 31 (2019). https://doi.org/10.1186/s40657-019-0171-8

